Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-03-11 ResponseLetter2.tex

Thanks to the authors for improving the connection with previous rescue theory. A few minor comments on the responses to my comments below. Line numbers in tracked changes version.

\textcolor{blue}{We thank the reviewer for their patience and thorough suggestions. At this point, they have basically taught us the evolutionary rescue literature. 
This was going above and beyond their responsibilities. 
We apologize that we didn't do this ourselves, and really appreciate the reviewer's efforts.}

lines 69-93: The review of multi-step rescue in the intro makes some helpful connections but the logical flow could be improved. Yes, Martin et al 2013 extended Orr $\&$ Unckless 2008 by allowing for multiple different rescue mutations to occur during wildtype decline, but for the purposes of this paper the more important (and yet unmentioned) extension is to allow rescue by multiple (and in particular, two) mutational steps. I think it is then logical to skip to Osmond et al 2020 and instead of focusing on the fitness landscape, which is irrelevant here, focus on the fact that they approximate the 2-step result of Martin et al 2013 in two regimes: sufficiently subcritical and sufficiently critical single mutants. Ignoring the fitness landscape, this is essentially the same result as in Iwasa et al 2003 and 2004, who were specifically thinking about drug resistance and host shifts. I know I mentioned the 2-locus Uecker and Hermisson (2016) result, as it is 2-step rescue, but without more explanation its not clear how it is relevant to the current paper and, in its current placement, breaks up the flow.

\textcolor{blue}{ We have tried to make these changes. The relevant paragraphs now read:}
\begin{displayquote}
	\textcolor{blue}{
    There has been extensive theoretical analysis of evolutionary rescue. Orr and Unckless (2008,
    2014) thoroughly investigated the simplest case in which there is a specific single mutation or low-
    frequency variant that i ssufficient to rescue the population. 
    Martin et al. (2013) extended these results
    to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in
    offspring number. 
    Importantly for the present work, they also argued that two-step processes involving
    a ``transient'' intermediate genotype with growth rate close to zero may be an important form of rescue;
    however, they only calculated that the rate of such rescues should scale with the total number of such
    transient genotypes produced. Osmond et al. (2020) derived approximate expressions for the rate of
    these two-step rescues in limiting cases and found that for some parameter values two-step rescue is
    more likely than rescue by a single step.}

    \textcolor{blue}{
    Resistance to cancer therapy is a natural application of evolutionary rescue theory (Alexander et al, 2014). 
    Iwasa et al (2003) used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction, focusing on the case in which the cancer must get two mutations to be rescued, with the single-mutant still dying rapidly.
    Iwasa et al (2004) used a similar approach to find simple approximate expressions for the probability that a single lineage would evolve to survive via a wide range of possible mutational pathways.
    These results are essentially the same as those described later in the general context by Osmond et al (2020) for the probability of evolutionary rescue by multiple mutations,
    which we mentioned in the previous paragraph\ldots}
\end{displayquote}

@@

\textcolor{blue}{We have made this change.}

lines 582-585: The same issue about a landscape arises here. Again, I don't think it is helpful to think of the current paper as Osmond et al 2020 with a different fitness lanscape. Relatedly, I find the sentence "Comparing our results to the evolutionary rescue literature, we complement the studies by Iwasa et al. (2003) and Osmond et al. (2020) on multi-step rescue by deriving simple approximations for a different fitness landscape" confusing, in part also because Iwasa et al 2003 say that their results hold for any fitness landscape. Why not just drop that sentence and move on to the 1 vs 2 step rescue result?

lines 585-586: "Similar to Osmond et al. (2020), we find that two-step rescues can be more likely than simple one-step rescues." I think it would be worth adding the reason, because of a higher mutation rate to stepping-stone vs rescue mutations.

\textcolor{blue}{We have made both these changes. The paragraph now begins, ``Similar to Osmond et al (2020), we find that two-step rescues can be more likely than simple one-step rescues, because the mutation rate to stepping-stone (aneuploid) genotypes can be much higher than that to fully resistant genotypes.''}

@@

lines 606-607: "Future work could test when these effects have significant effect on the results of our model." This sentence could be replaced with something more informative. I.e., one could hypothesize that rates of aneuploid loss are small enough relative to the fitness gain over the wildtype to ignore and that allowing a resistant euploid to have a higher fitness than a resistant aneuploid would decrease the importance of 2-step rescue (by some relatively small constant proportion?).

\textcolor{blue}{Agreed. We have rewritten this paragraph as follows:
\begin{displayquote}
    Our model neglects the possibility of back-mutations from aneuploidy to euploidy, and also the possibility that the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation. 
    Both of these would be expected to reduce the importance of two-step rescues via aneuploidy.
    But for the former, unless the rate of loss of extra chromosomes is extremely high (for estimates in yeast, see Hose et al (2024)), comparable in magnitude to the growth rate $r_a$, we expect that its effect on the dynamics will be negligible.
    The latter possibility, in which aneuploidy substantially reduces the mutant growth rate, seems more likely
    to have an effect. 
    Including it would be a straightforward extension to our model.
    In this case, it could actually be more important to consider the possibility of the loss of aneuploidy, 
    as one would need to check the relative rates of the simple \textit{sensitive} $\rightarrow$ \textit{euploid mutant} path and the three-step \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{aneuploid mutant} $\rightarrow$ \textit{euploid mutant} path (Kohanovski et al, 2024).
\end{displayquote}
}

### 2025-03-11 ms.tex
updated response to reviewer 2

We will address these questions essentially by combining known results from the evolutionary rescue literature, and considering the quantities and parameter values that are relevant for cancer.
Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, as in \citet{iwasa2003evolutionary, iwasa2004evolutionary, osmond_genetic_2020}.

@@

Comparing our results to the evolutionary rescue literature, we complement the studies by \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020}, re-deriving some of their results and extending them by including intermediate (aneuploid) genotypes with growth rates substantially above one. 
Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues, because the mutation rate to stepping-stone (aneuploid) genotypes can be much higher than that to fully resistant genotypes. 
The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of the drug, is particularly interesting.
Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumors under specific drug therapies.
Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as they may enhance the probability of rescue (\Cref{rescue_prob}A) and extend the window of opportunity for rescue (\Cref{cdffig}).

>>>>

We will address these questions essentially by combining known results from the evolutionary rescue literature, and considering the quantities and parameter values that are relevant for cancer.
Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, as in \citet{iwasa2003evolutionary, iwasa2004evolutionary} and \citet{osmond_genetic_2020}.

@@

Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues, because the mutation rate to stepping-stone (aneuploid) genotypes can be much higher than that to fully resistant genotypes. 
The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of the drug, is particularly interesting.
Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumors under specific drug therapies.
Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as they may enhance the probability of rescue (\Cref{rescue_prob}A) and extend the window of opportunity for rescue (\Cref{cdffig}).
